---
title: "Assignment3B"
author: "Cecilia De La Fuente Chavez"
format: html
execute:
  echo: true
editor: visual
---

## Approach Deliverable

### Overview

For Assignment 3B, I will use the [U.S. Energy Information Administration's](https://www.eia.gov/renewable/data.php) *Total energy consumption of renewable and nonrenewable energy sources* data set that I used for Assignment 1. Because this data set does not have daily points, only yearly and monthly data, I will calculate the year-to-date average and the six-month moving averages, instead of six-day, for two items, Solar and Wind energy consumption. I will use window functions in dplyr, because I prefer working in R.

## Code Base Deliverable

### Load library

```{r}
library(tidyverse)
```

### Read data from GitHub

Original file is a xlsx with two tabs. I saved the "Monthly Data" as a csv and stored it on GitHub, I used the skip argument in the read_csv() function to omit the first rows of metadata.

```{r}
url <- "https://raw.githubusercontent.com/ceciliadelafuentech/DATA607-Week3B/refs/heads/main/Table_1.3_Primary_Energy_Consumption_by_Source.csv"  

df_raw <- read_csv(   file = url,   
                  show_col_types = FALSE,   
                  progress = FALSE,    
                  skip = 10 
                  ) 
glimpse(df_raw)
```

### Cleaning and transformation process

I used row indexing to remove the first row after headers which was unnecessary for this analysis, and saved the updated data frame to a new one called df.

```{r}
df <- df_raw[-1,] 
glimpse(df)
```

Used mutate to create a new column, Date that takes the character format date from Month, and stores it as a date format.

```{r}
df <- df %>%
  mutate(
    Date = as.Date(paste0("01 ", Month), format = "%d %Y %B")
  )
glimpse(df)
```

For this assignment, I will only be doing a time series for Solar and Wind energy consumption. Because each of these is its own column, I am combining them as values into a new column called Source and their respective values into a another column called Consumption. This will allow me to more easily group by energy source for the analysis.

```{r}
df_long <- df %>%
  select(Date, `Solar Energy Consumption`, `Wind Energy Consumption`) %>%
  pivot_longer(
    cols = c(`Solar Energy Consumption`, `Wind Energy Consumption`),
    names_to = "Source",
    values_to = "Consumption"
  )
glimpse(df_long)
```

Change the values in Consumption column to numeric type.

```{r}
df_long <- df_long %>%
  mutate(Consumption = as.numeric(Consumption))
glimpse(df_long)
```

This code chunk arranges the data by energy source type and date, then groups it by year to calculate the cumulative average and store those values in a new column ytd_average. After, I used ungroup() to return the data frame to its normal structure.

```{r}
df_long <- df_long %>%
  arrange(Source, Date) %>%
  group_by(Source, Year = lubridate::year(Date)) %>%
  mutate(
    ytd_average = cummean(Consumption)
  ) %>%
  ungroup()
```

This code chunk shows the year-to-date average for solar energy consumption for each month in 2025.

```{r}
df_long %>%
  filter(Source == "Solar Energy Consumption", Year == 2025) %>%
  select(Date, Source, Consumption, ytd_average)
```

This code chunk arranges the data by energy source type and date, then groups it by Source to calculate the average for each six-month window, and store those values in a new column six_month_moving_average. After, I used ungroup() to return the data frame to its normal structure.

```{r}
df_long <- df_long %>%
  arrange(Source, Date) %>%
  group_by(Source) %>%
  mutate(
    six_month_moving_average = slider::slide_dbl(
      Consumption,
      mean,
      .before = 5,
      .complete = TRUE
    )
  ) %>%
  ungroup()
```

This code chunk shows the six-month moving average for wind energy consumption for each month in 2025.

```{r}
df_long %>%
  filter(Source == "Wind Energy Consumption", Year == 2025) %>%
  select(Date, Source, Consumption, six_month_moving_average)
```

### Visualizations

Solar Plot - create a subset of the data frame to make the a plot with Solar data from 2020 to 2026

```{r}
solar_plot <- df_long %>%
  filter(
    Source == "Solar Energy Consumption",
    Year >= 2020,
    Year <= 2026
  )
```

Line graph with 3 lines showing the Monthly consumption, the ytd average and the six-month average

```{r}
ggplot(solar_plot, aes(x = Date)) +
  geom_line(aes(y = Consumption, color = "Monthly Consumption"), alpha = 0.5) +
  geom_line(aes(y = ytd_average, color = "YTD Average")) +
  geom_line(aes(y = six_month_moving_average, color = "Six-Month Moving Average")) +
  labs(
    title = "Solar Energy Consumption: 2020–2026",
    x = "Date",
    y = "Consumption (Quadrillion Btu)"
  )
```

Wind Plot - create a subset of the data frame to make the a plot with Wind data from 2020 to 2026

```{r}
wind_plot <- df_long %>%   
  filter(     
    Source == "Wind Energy Consumption",     
    Year >= 2020,     
    Year <= 2026   
    )
```

Line graph with 3 lines showing the Monthly consumption, the ytd average and the six-month average

```{r}
ggplot(wind_plot, aes(x = Date)) +
  geom_line(aes(y = Consumption, color = "Monthly Consumption"), alpha = 0.5) +
  geom_line(aes(y = ytd_average, color = "YTD Average")) +
  geom_line(aes(y = six_month_moving_average, color = "Six-Month Moving Average")) +
  labs(
    title = "Wind Energy Consumption: 2020–2026",
    x = "Date",
    y = "Consumption (Quadrillion Btu)"
  )
```

### Conclusion

The window functions help identify yearly and six-month trends in solar and wind energy consumption. Monthly fluctuations can be observed, and an overall upwards trend from 2020 until 2026.
